---
title: "Simulation based approach for sample size in cluster RCT with binary endpoint"
subtitle: "`r knitr::current_input(dir = TRUE)`"
author: "Mark Jones"
date: "`r Sys.time()`"
output:
  html_document:
    number_sections: yes
    self_contained: yes
    theme: united
    toc: yes
    toc_depth: 3
geometry: left=0.2cm,right=0.2cm,top=1cm,bottom=1cm
editor_options:
  chunk_output_type: console
---

<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = 'figs/')
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(compiler))
suppressPackageStartupMessages(library(geepack))
suppressPackageStartupMessages(library(doParallel))
suppressPackageStartupMessages(library(foreach))
suppressPackageStartupMessages(library(beepr))
suppressPackageStartupMessages(library(clusterPower))

ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(text = element_text(size = 10))
ggplot2::theme_update(legend.position = "top")
#ggplot2::theme_update(legend.title = element_blank())
ggplot2::theme_update(axis.text.x = element_text(size = 10))
ggplot2::theme_update(axis.text.y = element_text(size = 10))

prob_to_odd <- cmpfun(function(x){
  return(x/(1-x))
})

inv_logit <- cmpfun(function(x){
  return(exp(x)/(1+exp(x)))
})

# Work|Right|Fast
# rmarkdown::render("simulation_report.Rmd", clean=TRUE)
```

# Data generation

Data model follows. $x_{ij}$ is a subject level effect e.g. age, $z_j$ is a group level variable, e.g. treatment. Cluster level random intercept.

\begin{aligned}
logit(p_{ij}) = \gamma_{00} + \gamma_{10} x_{ij} + \gamma_{01} z_j + u_{0j} + u_{1j} z_{j}
\end{aligned}

```{r}

create_base <- cmpfun(function(pars_id, rep_id, 
                               n_clusters, n_subjs_per_clust){
  
  stopifnot(n_clusters %% 2 == 0)
  ids <- 1:n_subjs_per_clust
  d0 <- data.table(pars_id = pars_id,
                   rep_id = rep_id,
                   id = rep(ids, len = length(ids) * n_clusters)) 
  d0[, clustid := rep(1:n_clusters, each = length(ids))]
  d0[, subjid := paste0(clustid, ":", id)]

  d0
})


gen_dat <- cmpfun(function(pars_id=1, rep_id=1, 
                           n_clusters = 4, 
                           n_subjs_per_clust = 5,
                           p0 = 0.30, 
                           gamma_10 = 0.3,
                           gamma_01 = 0.3,
                           sig_0 = 0.8){
  
  # static design matrix - same for all datasets
  d0 <- create_base(pars_id, rep_id, n_clusters, n_subjs_per_clust)
  n <- nrow(d0)
  
  # individual level covariate (centered and standardised)
  # could apply non-deterministic trt allocation
  d0[, x := rnorm(n)]
  d0[, z := rep(0:1, each = (n_clusters/2)*n_subjs_per_clust)]
  
  gen_clust <- function(x, sig_0){
    
    # hosp_cov <- 0.7*(sqrt(sig_0*sig_1))
    # vcov <- matrix(c(sig_0, hosp_cov, hosp_cov, sig_1), 2, 2)
    # clust_re <- MASS::mvrnorm(1, mu=c(0,0), Sigma = vcov)
    
    clust_re <- rnorm(1, 0, sig_0)

    data.table(clust_int = rep(clust_re, n_subjs_per_clust))
  }
  

  l <- lapply(1:n_clusters, gen_clust, sig_0=sig_0)
  d <- as.data.table(cbind(d0, rbindlist(l)))
  
  gamma_00 <- log(prob_to_odd(p0))
  
  d[, eta := gamma_00 + gamma_10*x + gamma_01*z + clust_int]
  d[, p := inv_logit(d$eta)]
  d[, y := rbinom(nrow(d), 1, prob = d$p)]

  # i should probably only return the stochastic parts
  d
})

```

Random effects $\sigma_0$ set to 0.35, 0.8, 1.5. In a fully unconditional model random intercept model this would yield ICCs of 0.04, 0.17 and 0.38.

```{r}
n_clusters <- 30
n_subjs_per_clust <- 10
d <- gen_dat(1, 1, n_clusters, n_subjs_per_clust, sig_0 = 0.35)
```

# Generate datasets

```{r}
replicate <- 1:500
num_clusters <- c(30, 50, 100)
people_per_cluster <- c(30, 50, 100)
sig_intercept <- c(0.2, 0.4, 0.8)
baseline_p0 <- c(0.1, 0.2, 0.3)
# treatment effect
log_or_z <- c(0.3, 0.6)

par_space <- data.table(expand.grid(replicate = replicate,
                                    num_clusters = num_clusters, 
                                    people_per_cluster = people_per_cluster,
                                    baseline_p0 = baseline_p0,
                                    sig_intercept = sig_intercept,
                                    log_or_z = log_or_z))

par_dim <- nrow(par_space)/max(replicate)
par_space[,pars_id := rep(1:par_dim, each = max(replicate))]   
setkey(par_space, pars_id, replicate)
```

Could pre-generate data like this and store.

```{r, eval = F}
gen_dat_wrap <- function(x){
  pars <- par_space[x,]
  
  gen_dat(pars_id= pars[,pars_id], 
          rep_id = pars[,replicate], 
          n_clusters = pars[,num_clusters], 
          n_subjs_per_clust = pars[,people_per_cluster], 
          p0 = pars[,baseline_p0], 
          gamma_01 = pars[,log_or_z], 
          sig_0 = pars[,sig_intercept])
}

set.seed(10001)
l <- lapply(1:nrow(par_space), gen_dat_wrap)
```

# Simulation
 
```{r, eval = F}
starttime <- Sys.time()

cl <- makeCluster(parallel::detectCores() - 2, outfile="")
registerDoParallel(cl)
# registerDoSEQ()

results <- foreach(i = 1:nrow(par_space),
                   .errorhandling = 'pass',
                   .export = c("prob_to_odd", "inv_logit", "par_space"),
                   .packages=c("MASS", "data.table", "compiler", "geepack")
                   #.options.snow=opts,
                   ) %dopar%{
 
  pars <- par_space[i,]
  d <- gen_dat(pars_id= pars[,pars_id], 
          rep_id = pars[,replicate], 
          n_clusters = pars[,num_clusters], 
          n_subjs_per_clust = pars[,people_per_cluster], 
          p0 = pars[,baseline_p0], 
          gamma_01 = pars[,log_or_z], 
          sig_0 = pars[,sig_intercept])    
   
  g1 <- geepack::geeglm(y ~ x + z, data = d, family = binomial,
                        id = clustid, corstr = "exchangeable")
  
  if(summary(g1)$coefficients["z", "Pr(>|W|)"]< 0.05){
    return(T)
  } else{
    return(F)
  }
  
}

stopCluster(cl)
endtime <- Sys.time()
difftime(endtime, starttime, units = "hours")
par_space[,win := unlist(results)]
```

Summarise results.

```{r}
par_space <- readRDS("results.RDS")
dpwr <- par_space[, .(pwr = 100*mean(win)), by = c("pars_id")]
dpwr <- merge(dpwr, unique(par_space[, .(pars_id, num_clusters,
                                         people_per_cluster, 
                                         baseline_p0, sig_intercept,
                                         log_or_z)]))



dpwr %>% dplyr::select(-pars_id) %>% 
  kable(caption = "Table 1. Power",
      col.names = c("Power", "Clusters", "Cluster size", "$p_0$", "$\\sigma_u$", "log odds ratio (trt)"),escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = F, position = "left",
                font_size = 12,
                latex_options = "hold_position")
# beepr::beep(6)
```




# Visualisation

```{r, fig.height=8, fig.width=7, echo = F}
dpwr[,icc := sig_intercept^2/(sig_intercept^2 + (pi^2)/3)]
dtmp <- dpwr %>%
  dplyr::filter(baseline_p0 == 0.1, sig_intercept == 0.2)

ggplot(data = dpwr, aes(x= num_clusters, 
                        y = pwr, 
                        colour = paste0(people_per_cluster),
                        lty = paste0(round(exp(log_or_z), 1)),
                        group = paste0(people_per_cluster, ":", 
                                       exp(log_or_z)))) +
  geom_point()+
  geom_line()+
  scale_linetype("Odds ratio for trt")+
  scale_color_discrete("Cluster size")+
  scale_y_continuous("Power",lim = c(0, 100))+
  scale_x_continuous("Number of clusters")+
  facet_grid(paste0("ICC ", round(icc, 3)) ~ 
               paste0("Baseline prev ", baseline_p0))
```

# ClusterPower

A few arbitrary points from cluster power package. A bit different, presumably because I had an additional covariate in the model.

```{r}
crtpwr.2prop(alpha = 0.05, power = NA, m = 50, n = 50, cv = 0,
  p1 = 0.13, p2 = 0.1, icc = 0.012, pooled = F, p1inc = T,
  tol = .Machine$double.eps^0.25)

crtpwr.2prop(alpha = 0.05, power = NA, m = 50, n = 100, cv = 0,
  p1 = 0.13, p2 = 0.1, icc = 0.012, pooled = F, p1inc = T,
  tol = .Machine$double.eps^0.25)

crtpwr.2prop(alpha = 0.05, power = NA, m = 30, n = 100, cv = 0,
  p1 = 0.25, p2 = 0.2, icc = 0.046, pooled = F, p1inc = T,
  tol = .Machine$double.eps^0.25)

crtpwr.2prop(alpha = 0.05, power = NA, m = 50, n = 30, cv = 0,
  p1 = 0.37, p2 = 0.3, icc = 0.163, pooled = F, p1inc = T,
  tol = .Machine$double.eps^0.25)
```




