---
title: "Simulation sample size in cluster RCT with binary endpoint and 3 arms"
subtitle: "`r knitr::current_input(dir = TRUE)`"
author: "Mark Jones"
date: "`r Sys.time()`"
output:
  html_document:
    number_sections: yes
    self_contained: yes
    theme: united
    toc: yes
    toc_depth: 3
geometry: left=0.2cm,right=0.2cm,top=1cm,bottom=1cm
editor_options:
  chunk_output_type: console
---

<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = 'figs/')
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(compiler))
suppressPackageStartupMessages(library(geepack))
suppressPackageStartupMessages(library(doParallel))
suppressPackageStartupMessages(library(foreach))
suppressPackageStartupMessages(library(beepr))
suppressPackageStartupMessages(library(clusterPower))
suppressPackageStartupMessages(library(randomizeR))
suppressPackageStartupMessages(library(multcomp))
suppressPackageStartupMessages(library(doBy))




ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(text = element_text(size = 10))
ggplot2::theme_update(legend.position = "top")
#ggplot2::theme_update(legend.title = element_blank())
ggplot2::theme_update(axis.text.x = element_text(size = 10))
ggplot2::theme_update(axis.text.y = element_text(size = 10))

prob_to_odd <- cmpfun(function(x){
  return(x/(1-x))
})

inv_logit <- cmpfun(function(x){
  return(exp(x)/(1+exp(x)))
})
odd <- cmpfun(function(p){p/(1-p)})
# Work|Right|Fast
# rmarkdown::render("simulation_report.Rmd", clean=TRUE)
```

# Data generation

Data model follows. Indicators for treatment group. Include SES and HEPC (hep-c clinic). Cluster level random intercept.

\begin{aligned}
logit(p_{ij}) = \beta_0 + \beta_1 \text{I(trt1)}_{ij} + \beta2 \text{I(trt2)}_{ij} + \beta_3 \text{SES}_{ij} + \beta_4 \text{hepc}_{ij} + \text{clust}_j
\end{aligned}

```{r}

create_base <- cmpfun(function(pars_id, rep_id, 
                               n_clusters, cluster_sizes){
  
  stopifnot(n_clusters == length(cluster_sizes))
  
  ids <- unlist(lapply(seq(1, length(cluster_sizes)), 
                       function(x) 1:cluster_sizes[x] ))
  d0 <- data.table(pars_id = pars_id,
                   rep_id = rep_id,
                   id = ids)
  clustids <- unlist(lapply(seq(1, n_clusters), 
                            function(x) rep(x, each = cluster_sizes[x]) ))
  d0[, clustid := clustids]
  d0[, subjid := paste0(clustid, ":", id)]
  d0
})


gen_dat <- cmpfun(function(pars_id=1, rep_id=1, 
                           n_clusters = 5, 
                           n_subjs_per_clust = 5,
                           n_subjs_per_clust_cv = 0.5,
                           p0 = 0.30, 
                           beta1 = 0.3,   # trt1
                           beta2 = 0.3,   # trt2
                           beta3 = 0.3,   # ses
                           beta4 = 0.3,   # hc serv
                           sig_0 = 0.8){
  
  # static design matrix - same for all datasets
  sig_siz <- n_subjs_per_clust * n_subjs_per_clust_cv
  # the plus 1 deals with zeros
  cluster_sizes <- abs(ceiling(rnorm(n_clusters, n_subjs_per_clust, sig_siz)))+1
  # dput(cluster_sizes)
  d0 <- create_base(pars_id, rep_id, n_clusters, cluster_sizes)
  n <- nrow(d0)
  stopifnot(n == sum(cluster_sizes))
  
  # individual level covariate (centered and standardised)
  # could apply non-deterministic trt allocation
  prac_ses <- unlist(lapply(seq(1, n_clusters), 
                            function(x) rep(sample(1:8, 1), 
                                            each = cluster_sizes[x]) ))
  hep_c_serv <- unlist(lapply(seq(1, n_clusters), 
                              function(x) rep(rbinom(1, 1, prob = 0.25), 
                                              each = cluster_sizes[x]) ))

  c <- as.numeric(genSeq(rpbrPar(n_clusters, 3, K = 3))@M)

  stopifnot(length(c) == n_clusters)
  stopifnot(length(cluster_sizes) == n_clusters)
  trt <- unlist(lapply(seq(1, n_clusters), function(x) rep(c[x], 
                                                           each = cluster_sizes[x]) ))

  stopifnot(length(trt) == nrow(d0))
  trtm <- model.matrix(~ factor(trt))
  
  d0[, prac_ses := prac_ses]
  d0[, hep_c_serv := hep_c_serv]
  d0[, trt := as.factor(trt)]
  d0[, trt1 := as.numeric(trtm[,2])]
  d0[, trt2 := as.numeric(trtm[,3])]

  clust_int <- unlist(lapply(seq(1, n_clusters), 
                             function(x) rep(rnorm(1, sd = sig_0), 
                                             each = cluster_sizes[x]) ))
  d0[, clust_int := clust_int]

  beta0 <- log(prob_to_odd(p0))

  d0[, eta := beta0 + beta1*trt1 + beta2*trt2 + beta3*prac_ses + beta4*hep_c_serv + clust_int]
  d0[, p := inv_logit(d0$eta)]
  d0[, y := rbinom(nrow(d0), 1, prob = d0$p)]

  # i should probably only return the stochastic parts
  d0
})
```

Random effects $\sigma_0$ set to 0.35, 0.8, 1.5. In a fully unconditional model random intercept model this would yield ICCs of 0.04, 0.17 and 0.38.

```{r}
iccme <- function(u){(u^2)/(u^2 + (pi^2)/3)}
sigme <- function(rho){ sqrt((rho*(pi^2)/3)/(1-rho))}
sig_u <- c(0.18, 0.42, 0.6)
iccme(sig_u)
```


```{r}
sig_u <- c(0.18, 0.42, 0.6)
n_clusters <- 30
n_subjs_per_clust <- 20
n_subjs_per_clust_cv <- 0.5
p0 <- 0.2
beta1 <- 0.5
beta2 <- 0.55

d <- gen_dat(1, 1, n_clusters, n_subjs_per_clust, 
             beta1 = beta1, beta2 = beta2, sig_0 = sig_u[1])
```

# Generate datasets

```{r}
replicate <- 1:500
n_clusters <- c(30, 50, 100)
n_subjs_per_clust <- c(30, 50, 100)
n_subjs_per_clust_cv = 0.5
sig_0 <- c(0.18, 0.42, 0.6)
p0 <- c(0.1, 0.2, 0.3)
# treatment effect
beta1 <- c(0.4, 0.5)
beta2 <- c(0.45, 0.55)

par_space <- data.table(expand.grid(replicate = replicate,
                                    n_clusters = n_clusters, 
                                    n_subjs_per_clust = n_subjs_per_clust,
                                    n_subjs_per_clust_cv = n_subjs_per_clust_cv,
                                    p0 = p0,
                                    beta1 = beta1,
                                    beta2 = beta2,
                                    sig_0 = sig_0))


par_dim <- nrow(par_space)/max(replicate)
par_space[,pars_id := rep(1:par_dim, each = max(replicate))]   
setkey(par_space, pars_id, replicate)


# now remove out any duplicates that we do not want
# e.g. we could remove the case where both treatments are equal
# par_space <- par_space[beta1 != beta2,]



```

Could pre-generate data like this and store.

```{r, eval = F}
gen_dat_wrap <- function(x){
  pars <- par_space[x,]
  
  gen_dat(pars_id=pars[,pars_id], 
               rep_id=pars[,replicate], 
               n_clusters = pars[,n_clusters], 
               n_subjs_per_clust = pars[,n_subjs_per_clust],
               n_subjs_per_clust_cv = pars[,n_subjs_per_clust_cv],
               p0 = pars[,p0],
               beta1 = pars[,beta1],   # trt1
               beta2 = pars[,beta2],   # trt2
               sig_0 = pars[,sig_0])
}

set.seed(10001)
l <- lapply(1:nrow(par_space), gen_dat_wrap)
```

# Simulation
 
```{r, eval = T}
starttime <- Sys.time()

cl <- makeCluster(parallel::detectCores() - 3, outfile="")
registerDoParallel(cl)
# registerDoSEQ()

results <- foreach(i = 1:nrow(par_space),
                   .errorhandling = 'pass',
                   .export = c("prob_to_odd", "inv_logit", "par_space"),
                   .packages=c("MASS", "data.table", "compiler", 
                               "geepack", "doBy", "randomizeR", "stats")
                   #.options.snow=opts,
                   ) %dopar%{
 
  pars <- par_space[i,]
  
  d <- gen_dat(pars_id=pars[,pars_id], 
               rep_id=pars[,replicate], 
               n_clusters = pars[,n_clusters], 
               n_subjs_per_clust = pars[,n_subjs_per_clust],
               n_subjs_per_clust_cv = pars[,n_subjs_per_clust_cv],
               p0 = pars[,p0],
               beta1 = pars[,beta1],   # trt1
               beta2 = pars[,beta2],   # trt2
               sig_0 = pars[,sig_0])
  
  # d0[, eta := beta0 + beta1*trt1 + beta2*trt2 + beta3*prac_ses + beta4*hep_c_serv + clust_int]
  
  g1 <- geepack::geeglm(y ~ trt + prac_ses + hep_c_serv, 
                        data = d, family = binomial,
                        id = clustid, corstr = "exchangeable")
  
  L <- matrix(c(-1,1,0,0,0,
                -1,0,1,0,0,
                0,-1,1,0,0), nrow = 3, byrow = T)
  p <- p.adjust(esticon(g1, L=L)[,"Pr(>|X^2|)"], method = "bonferroni")

  if(all(p < 0.05)){
    return(T)
  } else{
    return(F)
  }
  
}

stopCluster(cl)
endtime <- Sys.time()
difftime(endtime, starttime, units = "hours")
par_space[,win := unlist(results)]
```

Summarise results.

```{r}
#par_space <- readRDS("results.RDS")
dpwr <- par_space[, .(pwr = 100*mean(win)), by = c("pars_id")]
dpwr <- merge(dpwr, unique(par_space[, .(pars_id, n_clusters,
                                         n_subjs_per_clust, 
                                         n_subjs_per_clust_cv, 
                                         p0, beta1, beta2, sig_0
                                         )]))


dpwr %>% dplyr::select(-pars_id) %>% 
  kable(caption = "Table 1. Power",
      col.names = c("Power", "Clusters", "Cluster size", "Size CV", 
                    "$p_0$", "$\beta_1$","$\beta_2$", "$\\sigma_u$"),escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = F, position = "left",
                font_size = 12,
                latex_options = "hold_position")
# beepr::beep(6)
```




# Visualisation

The legend shows cluster size (coefficient of variation) as colour. Linetype shows the odds ratios for the first treatment group (e.g. funding to practice) and the second treatment group (e.g. funding to practice and participant) relative to the control condition.

```{r, fig.height=8, fig.width=7, echo = F}
dpwr[,icc := sig_0^2/(sig_0^2 + (pi^2)/3)]


ggplot(data = dpwr, aes(x= n_clusters, 
                        y = pwr, 
                        colour = paste0(n_subjs_per_clust, " (",
                                       n_subjs_per_clust_cv, ")",
                                       sep = "" ),
                        lty = sprintf("%.3f, %.3f", 
                                      exp(beta1), exp(beta2)),
                        group = paste(n_subjs_per_clust, 
                                       n_subjs_per_clust_cv, 
                                       beta1,  
                                       beta2, 
                                       sep = ", " )   )) +
  geom_point()+
  geom_line()+
  scale_linetype("ORs (trt)")+
  scale_color_discrete("Cluster size")+
  scale_y_continuous("Power",lim = c(0, 100))+
  scale_x_continuous("Number of clusters")+
  facet_grid(paste0("ICC ", round(icc, 3)) ~ 
               paste0("Baseline prev ", p0))+
  guides(colour=guide_legend(nrow=2,byrow=TRUE))+
  guides(lty=guide_legend(nrow=2,byrow=TRUE))
```

# ClusterPower

A few arbitrary points from cluster power package. A bit different, presumably because I had an additional covariate in the model.

```{r}
crtpwr.2prop(alpha = 0.05, power = NA, m = 50, n = 50, cv = 0,
  p1 = 0.13, p2 = 0.1, icc = 0.012, pooled = F, p1inc = T,
  tol = .Machine$double.eps^0.25)

crtpwr.2prop(alpha = 0.05, power = NA, m = 50, n = 100, cv = 0,
  p1 = 0.13, p2 = 0.1, icc = 0.012, pooled = F, p1inc = T,
  tol = .Machine$double.eps^0.25)

crtpwr.2prop(alpha = 0.05, power = NA, m = 30, n = 100, cv = 0,
  p1 = 0.25, p2 = 0.2, icc = 0.046, pooled = F, p1inc = T,
  tol = .Machine$double.eps^0.25)

crtpwr.2prop(alpha = 0.05, power = NA, m = 50, n = 30, cv = 0,
  p1 = 0.37, p2 = 0.3, icc = 0.163, pooled = F, p1inc = T,
  tol = .Machine$double.eps^0.25)
```



```{r, eval = F, echo = F}

inv_logit(log(prob_to_odd(0.1))+0.3)

```



