---
title: "Proportional hazards model - weibull"
subtitle: "`r knitr::current_input(dir = TRUE)`"
author: "Mark Jones"
date: "`r Sys.time()`"
output:
  html_document:
    number_sections: yes
    self_contained: yes
    theme: united
    toc: yes
    toc_depth: 3
geometry: left=0.2cm,right=0.2cm,top=1cm,bottom=1cm
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(rstan))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(nphsim))
suppressPackageStartupMessages(library(survival))
suppressPackageStartupMessages(library(flexsurv))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(brms))
suppressPackageStartupMessages(library(bayesplot))
```

# Preamble

Data generation using piecewise exponential specified with three baseline hazards.

```{r, echo = T}
mypwexp <- function(n, rates, intervals){
  
  k <- length(rates)
  tx <- 0
  j <- 1
  times <- array(0,n)
  timex <- cumsum(intervals)
  indx <- array(TRUE,n)
  for(i in 1:k){
    nindx <- sum(indx)
    if (nindx==0) break
    increment <- rexp(nindx,rates[i])
    times[indx] <- tx + increment
    if (i<k){
      tx <- timex[i]
      indx <- (times > timex[i])
    }
  }
  return(times)
  
}

gendat <- function(){
  rates <- c(0.05, 0.04, 0.03, 0.03, 0.02)
  intervals <- c(6, 12, 24, 36)
  n <- 100
  x0 <- mypwexp(n/2, rates, intervals)
  x1 <- mypwexp(n/2, rates - 0.015, intervals)
  d <- data.table(cen = rep(0, n), 
                  time = c(x0, x1),
                  grp = rep(0:1, each = n/2))
  d[, cen := as.numeric(d$time > 40)]
  d[cen == 1, time := 40]
  d[, event := 1-cen]
  
  # long format - don't need this, it is just for ref
  # split at each timepoint 
  d2 <- survSplit(data = d, 
                  cut = intervals, 
                  episode = "intvl",
                  start="start", 
                  id = "id", 
                  end="time", 
                  event="event")
  
  d2 <- d2 %>%
    dplyr::mutate(exposure = time - start,
                  intvl = factor(intvl,  
                             labels = paste("_", c(0,intervals), "_", 
                                            c(intervals,1000), sep="")))
  d
}
d <- gendat()
```

KM plot. 

```{r, echo = F}
fit <- survfit(Surv(time, event) ~ grp, data = d)
plot(fit, lty = 1:2,col=1:2)
legend(
  "topright",
  legend=c("PWE0", "PWE1"),
  col=1:2,
  lty=1:2,
  horiz=FALSE,
  bty='n')
dkmfig <- data.table(t = fit$time,
                     surv = fit$surv,
                     grp = c(rep(0, fit$strata[1]), rep(1, fit$strata[2])))
```


# Modelling

## ph model

If the hazard function factorises such that the covariates have a multiplicative effect on the hazard function then we have a PH model. It is common to use $exp(x_j \beta)$.

\begin{aligned}
h(t_j) &= h_0(t)g(x_j) \\
 &= h_0(t) exp(x_j \beta)
\end{aligned}

`flexsurv` is a much more sensible package than `survival`. It provides a PH parameterisation for weibull.

The coefficient for the `grp` estimate is negative which implies the risk of an event is decreased in the treatment group. The exponentiated parameter gives the multiplicative difference between the hazards in the two groups.

```{r}

# different parameterisation as per rweibull (affects shape and scale).
fs1 <- flexsurvreg(Surv(time, event) ~ grp, dist = "weibullPH", data = d)
print(fs1)

# the positive coefficient for grp imply longer time to event
# in the treatment group 
# the tte in the trt group is stretched by a factor of 1.54
exp(coef(fs1))
```


## stan - ph formulation

The pdf for weibull in stan is:

\begin{aligned}
f(t) = (a/b)(t/b)^{a-1}exp(-(t/b)^a)
\end{aligned}

The average survival time increases with an increasing values of the scale (b) parameter. So, the scale parameter is the one that is modelled with covariates. Typically like this:

\begin{aligned}
b_i = exp(-\frac{\mu + X_i'\beta}{b})
\end{aligned}

A positive $\beta$ means a shorter average survival time with a unit increase in $X$. 

```{r}
# these options help stan run quicker
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
d <- gendat()
sdat <- list(N = nrow(d), 
             M = 1,
             X = array(d$grp, dim = c(nrow(d), 1)), 
             y = d$time,
             event = d$event)

# Translate the Stan program to C++ code
c1 <- rstan::stanc(file = "003_weibull_ph_mod2.stan") 
# Construct the model from which to draw samples.
m1 <- rstan::stan_model(file = "003_weibull_ph_mod2.stan", verbose = F) 
# Might give the dreaded "There were 5 divergent transitions after warmup" wanring.
f2 <- rstan::sampling(m1, data = sdat, iter = 4000, 
                      chains = 1, thin = 1,
                      #seed = 99,
                      control = list(adapt_delta = 0.99))

print(f2, digits = 3)
fs1 <- flexsurvreg(Surv(time, event) ~ grp, dist = "weibullPH", data = d)
print(fs1)

# the intercept from the stan model should be about log(scale) from
# the flexsurvreg model. while the output from flexsurv shows the 
# exponentiated scale the object actually stores it without 
# exponentiating so stan model intercept should be about:
coef(fs1)["scale"]
```

## diagnostics

`bayesplot`  nuts diagnostics need more information than simply the posterior they need the log of the posterior density for each draw and some NUTS-specific diagnostic values.

```{r}
post <- as.array(f2)
# post is a multidim array.

lp <- log_posterior(f2)
mpars <- nuts_params(f2)

# https://cran.r-project.org/web/packages/bayesplot/vignettes/visual-mcmc-diagnostics.html#divergent-transitions
mcmc_parcoord(post, pars = c("intercept", "beta[1]", "shape"), np = mpars)

pairs(f2, pars = c("intercept", "beta[1]", "shape"), las = 1)

mcmc_scatter(post, pars = c("intercept", "shape"), 
  #transform = list(alpha = "log"), # can abbrev. 'transformations'
  np = mpars
)
```

## parameter estimates

```{r, fig.width = 4, fig.height=4}
mcmc_intervals(post, pars = c("beta[1]", "intercept"))
```

```{r, fig.width = 4, fig.height=4}
mcmc_hist(post, pars = c("beta[1]", "intercept"))
```

```{r, fig.width = 4, fig.height=4}
mcmc_trace(post, pars = c("beta[1]", "intercept"))
```

### survival function

I thought that the survival function for the stan parameterisation is $exp(-(\frac{t}{b})^a)$ where $a$ is the shape and $b$ is the scale. However, when you use $b = exp(\mu/shape)$ where $\mu = X\beta$ you actually have a survival function: $S(t) = exp(-b t^a)$. Why this is, I do not know and haven't been able to figure out.

```{r}
post <- as.matrix(f2)

wsurv_mean <- function(x, g, FUN = "mean", ...){
  t =  mtime[x]
  mu <- post[,"intercept"] + g * post[,"beta[1]"]
  shape <- post[, "shape"]
  # 
  scale <- exp(mu)
  match.fun(FUN)(exp(-scale* t^shape))
}

mtime <- seq(0, 100, len = 100)  
s0 <- unlist(lapply(1:length(mtime), wsurv_mean, g = 0))
s1 <- unlist(lapply(1:length(mtime), wsurv_mean, g = 1))
  
dfig <- data.frame(t = rep(mtime, len = 2 * length(mtime)),
                   grp = rep(0:1, each = length(mtime)),
                   surv = c(s0, s1))  

```



Dots show the KM estimates.

```{r, echo = F, fig.height = 5, fig.width = 5}
ggplot(dfig, aes(x = t, y = surv, group = grp, colour = paste0(grp)))+
  #geom_point(size = 0.3, alpha = 1) +
  geom_line()+
  geom_point(data = dkmfig,
             aes(x = t, y = surv, group = grp, colour = paste0(grp)),
             inherit.aes = F) +
  scale_color_discrete("Group") +
  scale_y_continuous(lim=c(0, 1))
```




### posterior things

work with the predictions as per weibull aft to example model parameter estimates from model fitted to new data.

```{r}

```



