---
title: "ICC in stepped wedge with binary response"
subtitle: "`r knitr::current_input(dir = TRUE)`"
author: "Mark Jones"
date: "`r Sys.time()`"
output:
  html_document:
    classoption: landscape
    number_sections: yes
    self_contained: yes
    theme: united
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: '3'
geometry: left=0.2cm,right=0.2cm,top=1cm,bottom=1cm
editor_options:
  chunk_output_type: console
classoption: landscape
---

<!--    toc: yes
    toc_float: true -->

<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>

# Preamble

## Random intercept model

To generate data for a cross-sectional SW design with a known ICC you can generate the baseline cluster proportions from a beta distribution, $p_{0k} \sim Beta(a, b)$. To do this solve:

\begin{aligned}
\mu &= \frac{a}{a+b} \\
\rho &= \frac{1}{a+b+1}
\end{aligned}

to get:

\begin{aligned}
b &= \frac{(1-\rho)(1-\mu)}{\rho} \\
a &= \frac{\mu(1-\rho)}{\rho}
\end{aligned}

the linear predictor is $logit(p_{jk}) = \beta_{0k} + \beta_1 X_{jk}$ with X as treatment indicator in cluster k at time k and $\beta_{0k} = log(\frac{p_{0k}}{1-p_{0k}})$.

In a stepped wedge design, if you have 30 clusters and 15 time points with each time point being an independent set of 50 people you get.

```{r}
suppressPackageStartupMessages(library(lme4))
a <- function(rho, mu)  (1-rho)*mu/rho
b <- function(rho, mu)  (1-rho)*(1-mu)/rho
odd <- function(p) p/(1-p)
inv_logit <- function(x) exp(x)/(1+exp(x))

n_clust <- 30
n_time <- 15
n_sub <- 50
rho <- 0.05
mu <- 0.6
beta1 <- 0.3
beta2 <- 0.03  
  
dgp1 <- function(){
  dstart <- data.frame(clust_id = 1:n_clust,
                     tx_start = rep(1:n_time, each = n_clust/n_time))

  d <- data.frame(expand.grid(clust_id = 1:n_clust,
                   t = 0:n_time,
                   sub_id = 1:n_sub), stringsAsFactors = F)
  
  d <- merge(d, dstart, by = "clust_id")
  d$tx_active <- as.numeric(d$t >= d$tx_start)
  # just to get unique sub ids
  d$sub_id_full <- paste("k", d$clust_id, "t", d$t, "i", d$sub_id, sep = "")
  d <- d[order(d$clust_id, d$t, d$sub_id),]
  # beta binomial
  d$p0 <- rep(rbeta(n_clust, shape1 = a(rho, mu), shape2 = b(rho, mu)),
            each = (n_time +1 ) * n_sub)
  
  d$eta <- log(odd(d$p0)) + beta1 * d$tx_active + beta2 * d$t
  d$p <- inv_logit(d$eta)
  d$y <- rbinom(nrow(d), 1, prob = d$p)
  d
}

d <- dgp1()

g1 <- geepack::geeglm(y ~ tx_active + t, family = binomial, data = d,
                      id = clust_id)
summary(g1)
l1 <- lme4::glmer(y ~ tx_active + t + (1|clust_id), family = binomial, data = d)
summary(l1)
sjstats::icc(l1)
```

The pre-specified ICC should be recovered along with the parameters. Note the biased estimate from the GEE due to small number of clusters. 

## Multilevel (3-level) model

Think clusters with patients with repeat measures.

According to Barker (Statistical considerations for estimating power of stepped wedge cluster randomised trials with cohorts of participants and a binary outcome) if you assume that $Y_{ijk}$ (i=subject, j= time, k=cluster) comes from some latent continuous measure a threshold value results in a 1 for the binary outcome then the level 1 variation can be assumed to come from a standard logistic distribution with mean zero and variance $\frac{\pi^2}{3}$. 

\begin{aligned}
\rho_3 &= \frac{\sigma^2_k}{\sigma_k^2 + \sigma_{i(k)}^2+ \frac{\pi^2}{3}} \\
\rho_2 &= \frac{\sigma_k^2}{\sigma_k^2 + \sigma_{i(k)}^2} \\
\rho_{23} &= \frac{\sigma_k^2 + \sigma_{i(k)}^2}{\sigma_k^2 + \sigma_{i(k)}^2 + \frac{\pi^2}{3}} \\
\end{aligned}

+ $\rho_3$ is the proportion of total variance which due to between cluster variance
+ $\rho_2$ is the correlation between participants within a cluster
+ $\rho_{23}$ is the correlation between measurements for the same participant

Solving these is tedious but eventually gives the following. Let's set $\rho_3 = 0.05$ and $\rho_2 = 0.5$. 

```{r}
# cluster level variance
var_k <- function(rho2, rho3){
  ((pi^2)/3) * (rho3*rho2/(rho2-rho3))
}
# subject level
var_i <- function(rho2, rho3){
  ((pi^2)/3) * (rho3*rho2/(rho2-rho3)) * ((1/rho2)-1)
}

rho2 <- 0.5
rho3 <- 0.05
vark <- var_k(rho2, rho3)
vari <- var_i(rho2, vark)

c(vari, vark)
```

The DGP function needs a minor modification.

```{r}
n_clust <- 30
n_time <- 15
n_sub <- 3 # this time we assume repeat measures on the 50
rho <- 0.05
mu <- 0.6
beta1 <- 0.3
beta2 <- 0.03  
p0 <- 0.5
  
dgp2 <- function(){
  dstart <- data.frame(clust_id = 1:n_clust,
                     tx_start = rep(1:n_time, each = n_clust/n_time))

  d <- data.frame(expand.grid(clust_id = 1:n_clust,
                   t = 0:n_time,
                   sub_id = 1:n_sub), stringsAsFactors = F)
  
  d <- merge(d, dstart, by = "clust_id")
  d$tx_active <- as.numeric(d$t >= d$tx_start)
  # just to get unique sub ids
  d$sub_id_full <- paste("k", d$clust_id, "i", d$sub_id, sep = "")
  
  dclustint <- data.frame(clust_id = 1:n_clust,
                          clust_int = rnorm(n_clust, 0, vark))
  
  dsubint <- data.frame(expand.grid(clust_id = 1:n_clust,
                                    sub_id = 1:n_sub))
  dsubint <- dsubint[order(dsubint$clust_id, dsubint$sub_id),]
  dsubint$sub_int <- rnorm(nrow(dsubint), 0, vari)
  
  d <- merge(d, dclustint, by = "clust_id")
  d <- merge(d, dsubint, by =c("clust_id", "sub_id"))
  d <- d[order(d$clust_id, d$sub_id, d$t),]
  
  
  d$eta <- log(odd(p0)) + beta1 * d$tx_active + beta2 * d$t + d$clust_int + d$sub_int
  d$p <- inv_logit(d$eta)
  d$y <- rbinom(nrow(d), 1, prob = d$p)
  d
}

# it's rare to be able to recover 
set.seed(5)
d <- dgp2()
l1 <- lme4::glmer(y ~ tx_active + t + (1|clust_id/sub_id), family = binomial, data = d)
summary(l1)

# I don't trust this:
sjstats::icc(l1, adjust = F)

# but I think that computing is actually not this straight forward anyway.
myicc <- function(sig_top, sig_total){
  sig_top/sig_total
}
sig <- VarCorr(l1)
varsub <- as.numeric(sig[1])
varclust <- as.numeric(sig[2])
myicc(varsub, (varsub + varclust + ((pi^2)/3)))
myicc(varclust, (varsub + varclust + ((pi^2)/3)))
```

# Summary

Ran through the DGP to simulate data with pre-specified ICC values for both 2-level and 3-level hierarchical models with a stepped-wedge design.



